---
title: "Introduction to kenyacensus"
author: "Kenya Census Data Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to kenyacensus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `kenyacensus` package provides comprehensive Kenya Population and Housing Census data from 1948 to 2019. All data is sourced from the Kenya National Bureau of Statistics (KNBS).

## Installation

```r
# Install from source
devtools::install_github("kenya-data/kenyacensus")

# Or install from local directory
devtools::install("path/to/kenyacensus")
```

## Loading the Package

```r
library(kenyacensus)
```

## Listing Available Datasets

```{r}
list_datasets()
```

## National Trends

View population growth over time:

```{r}
head(census_age_sex)
```

Calculate growth rates:

```{r}
calculate_growth_rate(2009, 2019)
```

Plot the trend:

```{r, fig.width=7, fig.height=5}
plot_census_trend()
```

## County-Level Analysis (2019)

Get data for specific counties:

```{r}
get_county_data(c("Nairobi City", "Mombasa", "Kisumu"))
```

Find the most populous counties:

```{r}
head(census_by_county[order(-census_by_county$Total), c("County", "Total", "Population_Density")], 10)
```

Compare counties:

```{r}
compare_counties(c("Nairobi City", "Mombasa", "Kisumu", "Nakuru"))
```

Search for counties:

```{r}
search_counties("Nai")
```

## Demographics

### Population Pyramid

```{r, fig.width=7, fig.height=6}
plot_population_pyramid()
```

### Age Structure

```{r}
census_age_sex
```

### Ethnicity

```{r}
head(census_ethnicity[order(-census_ethnicity$Population), ], 10)
```

## Education

```{r}
census_education
```

## Housing and Services

### Water Access

```{r}
census_water
```

### Lighting

```{r}
census_lighting
```

### Sanitation

```{r}
census_sanitation
```

## ICT Access

```{r}
census_ict
```

## Employment

```{r}
census_employment
```

## Historical Data by Province

```{r}
# Rift Valley population over time
rv_data <- census_by_province[census_by_province$Province == "Rift Valley", ]
rv_data[, c("Year", "Total")]
```

## Data Sources

All data is sourced from:
- Kenya National Bureau of Statistics (KNBS): https://www.knbs.or.ke/
- OpenAfrica (CKAN): https://open.africa/

## Citation

If you use this package in your research, please cite the Kenya National Bureau of Statistics as the original data source.
